---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Part 1:Loading and Preprocessing the Data
```{r}
        
        fileUrl<-"http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        download.file(url=fileUrl,destfile="Activity.zip")
        unzip("Activity.zip")
        activity<-read.csv("activity.csv")
        class(activity)
        str(activity)
        head(activity)
        tail(activity)
        summary(activity)
```

### Examining Missing Values
```{r}
                sum(is.na(activity$steps))
```


### Missing Values by Date
```{r}          
                missing_values<-tapply(is.na(activity$steps),activity$date,sum)
                missing_values[missing_values>0]
```
        
### Distribution of steps 
```{r}
                quantile(activity$steps,na.rm=TRUE,probs=c(0.5,0.6,0.7,0.8,0.9,0.95,1))
```




## Part 2:Mean and Median Total Sumber of Steps Taken Per Day
### Histogram of Total Number of Steps Taken Per Day
```{r,fig.height=4,fig.width=9}
                total_steps<-tapply(activity$steps,activity$date,sum,na.rm=TRUE)
                summary(total_steps)
                hist(total_steps,xlab="Total steps per day",
                     main="Histogram of total daily steps")
```

### Computing Mean and Median Total Steps Taken Per Day
```{r}
                mean_total_steps<-mean(total_steps)
                mean_total_steps
                median_total_steps<-median(total_steps)
                median_total_steps
```




## Part 3:Average Daily Activity Pattern
### Number of steps for each 5 minute interval averaged across all days
```{r}
                interval_steps<-tapply(activity$steps,as.factor(activity$interval)
                                       ,mean,na.rm=TRUE)
```
### Median number of steps for each 5 minute interval across all days
```{r}
                interval_steps_median<-tapply(activity$steps,as.factor(activity$interval),
                              median,na.rm=TRUE)
```
###Time Series Plot of 5-minute interval versus Average steps 
```{r,fig.height=4,fig.width=9}
                uniq_intervals<-unique(activity$interval)
                plot(uniq_intervals,interval_steps,type="l",xlab="Interval",
                     ylab="Average Number of Steps")
                abline(v=max_interval,col="blue2")
```


### 5-minute interval containing maximum number of steps
```{r}
                max(interval_steps)
                max_interval<-uniq_intervals[interval_steps==max(interval_steps)]
                max_interval
```
       
#### Conclusion: The 835 interval contains the maximum activity. The average maximum activity 
#### for this time interval is 206.1698 steps
        


## Part 4:Imputing Missing Values
### Total number of Missing Values
```{r}
                missing_vals<-sum(is.na(activity))
                missing_vals
                isnas<-which(is.na(activity$steps))
```
### Strategy 1:Creating a new dataset by Replacing missing values with Mean of 5-minute interval
```{r}
                activity$impsteps<-replace(activity$steps,isnas,interval_steps)
```
### Strategy 2:Creating a new dataset by Replacing missing values with Median of 5-minute interval
```{r}
                activity$impsteps_median<-replace(activity$steps,isnas,interval_steps_median)
```
### Strategy 1: Analysis with mean imputed values
```{r}
                total_impsteps<-tapply(activity$impsteps,activity$date,sum)
                mean_total_impsteps<-mean(total_impsteps)
                mean_total_impsteps
                median_total_impsteps<-median(total_impsteps)
                median_total_impsteps
```
####Histogram of total number of steps for Strategy 1
```{r,fig.height=4,fig.width=9}
                hist(total_impsteps,xlab="Total steps per day",
                        main="Histogram of total daily steps after imputing mean values")
```


### Strategy 1: Comparison of means and medians before and after imputing mean missing values
####Mean Total Steps (without imputing missing values)= 9354.23
####Mean Total Steps (after imputing average values) =  10766.19
####Median Total Steps (without imputing missing values)= 10395
####Median Total Steps (after imputing average values) = 10766.19
                

### Strategy 2: Analysis with median imputed values
```{r}
                total_impsteps_median<-tapply(activity$impsteps_median,activity$date,sum)
                mean_total_impsteps_median<-mean(total_impsteps_median)
                mean_total_impsteps_median
                median_total_impsteps_median<-median(total_impsteps_median)
                median_total_impsteps_median
```
####Histogram of total number of steps for Strategy 2
```{r,fig.height=4,fig.width=9}
                        hist(total_impsteps_median,xlab="Total steps per day",
                        main="Histogram of total daily steps after imputing median values")
```

### Strategy 2: Comparison of means and medians before and after imputing median missing values
####Mean Total Steps without imputing missing values= 9354.23
####Mean Total Steps after imputing median values) =  9503.869 
####Median Total Steps without imputing missing values)= 10395
####Median Total Steps after imputing median values) = 10395    
### Impact of imputing Missing Values in Strategy 1 and Strategy 2

### Comparison between Strategy 1 and Strategy 2
#### Strategy 1 : Mean value increases, Median also increases
#### Strategy 2 : Mean value decreases, Median value remains the same
#### Since after imputing median values the histogram of the total number of steps more closely
####resembles the original histogram and estimates are less biased, Strategy 2 is preferred. 
                     


## Part 5: Differences in Activity patterns between weekdays and weekends
#### Since Strategy 2 appears to create less biases in the computation of mean and median values 
####I shall use the dataset created by following Strategy 2 (median imputation)
####for this part of the analysis    

### Determining day of week
```{r}
                activity$asdate<-as.Date(activity$date,"%Y-%m-%d")
                class(activity$asdate)
                activity$day<-weekdays(activity$asdate)
```
### Creating a  Binary factor variable for weekday or weekend
```{r}
                activity$wkend<-activity$day %in% c("Saturday","Sunday")
```
### Computing the Interval averages for weekdays and weekends
```{r}
                wkdayavg<-tapply(activity$impsteps_median[activity$wkend=="FALSE"],as.factor
                 (activity$interval[activity$wkend=="FALSE"]),mean)
                wkendavg<-tapply(activity$impsteps_median[activity$wkend=="TRUE"],
                 as.factor(activity$interval[activity$wkend=="TRUE"]),mean)
```
### Plotting Weekday versus Weekend activity
```{r,fig.height=7,fig.width=9}
                par(mfrow=c(2,1),mar=c(4,4,2,1))
                plot(uniq_intervals,wkdayavg,type="l",col="blue",
                main=list("Weekdays",cex=1),xlab="",ylab="Number of steps")        
                plot(uniq_intervals,wkendavg,type="l",col="violet",
                main=list("Weekends",cex=1),xlab=list("Interval",cex=1.25),
                ylab="Number of steps")
```

### Conclusions: 
####There seem to be considerable differences between weekdays and weekends
#### 1. The maximum number of steps during weekdays is higher than in the weekends
#### 2. The number of steps during weekdays spikes sharply in the morning hours around  8 to 9 am 
#### 2. The average number of steps is distributed more uniformly across the various 
#### 5-minute intervals in the weekend than in the weekdays.
